\pkg{sdmTMB}: An \proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields

### 2023-09-26 doc/paper-jss/sdmTMB-paper.Rnw
Add todo
Wrap by sentence; some edits

Data are often collected in space or in space repeatedly over time. While such data are a rich source of information \citep[e.g.,][]{legendre1989a, rossi1992, tilman1997spatial}, they are challenging to properly model---data closer in space and time are usually more similar to each other than data farther apart due to measured and unmeasured variables \citep{cressie1993, diggle2007, cressie2011}. While measured variables can be accounted for with predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables (e.g., everything influencing species abundance but not explicitly modelled) can cause residual spatial correlation. Accounting for this residual correlation is important because doing so allows for valid statistical inference \citep{legendre1989a, dormann2007}, can improve predictions \citep[e.g.,][]{shelton2014}, and can be of interest itself \citep[e.g.,][]{thorson2019d, barnett2021}.

Geostatistical generalized linear mixed effects models (GLMMs) with spatially correlated random effects are a class of models appropriate for these data \citep{rue2005gmrf, diggle2007, cressie2011}. Similarly to how random intercepts can account for correlation among groups, spatial or spatiotemporal random effects can account for unmeasured variables causing observations to be correlated in space or space and time. A common approach to modelling these spatial effects is with Gaussian random fields (GRFs), where the random effects describing the spatial patterning are assumed to be drawn from a multivariate normal (MVN) distribution, constrained by some covariance function such as the exponential or Matérn \citep{cressie1993, diggle2007, chiles1999}.
% TODO: decide on using GRF throughout

Such models quickly become computationally limiting due to the need to invert large matrices to keep track of covariation among data \citep[e.g.,][]{rue2005gmrf, latimer2009}. Many solutions have been proposed, such as predictive processes \citep{banerjee2008, latimer2009}, the stochastic partial differential equation (SPDE) approximation to GRFs \citep{lindgren2011}, and nearest-neighbour Gaussian processes \citep{datta2016, finley2022}. These approaches reduce the scale of the covariance estimation problem while providing a means to evaluate the data likelihood, thereby allowing fitting via Bayesian \citep{gelfand2017} or maximum likelihood methods. This can greatly improve computational efficiency \citep[e.g.,][]{heaton2019}. The SPDE approach is a solution popularized via the \proglang{R}-\pkg{INLA} package \citep{rue2009, lindgren2011, lindgren2015} and an implementation in \pkg{TMB} \citep[Template Model Builder,][]{kristensen2016} that relies on \proglang{R}-\pkg{INLA} to create input matrices \citep[e.g.,][]{thorson2019, osgood-zimmerman2021}.
% Details are beyond the scope of this paper and are not necessary to use the software discussed here, but the idea is that the solution to a specific SPDE is a GRF with a Matérn covariance function and this `trick' enables one to efficiently fit approximations to GRFs to large spatial datasets \citep{lindgren2011}.

@@

\sdmTMB\ relies on several well-established \R\ packages (Figure~\ref{fig:flowchart}).
\sdmTMB\ uses \pkg{fmesher} \citep{fmesher} and \INLA\ \citep{lindgren2015} to generate the triangulation mesh and input matrices for the SPDE calculations.
The SPDE precision matrix calculations themselves are performed by functions within \pkg{TMB} \citep{kristensen2016} (\texttt{R\_inla} namespace).
Data structures required to implement penalized smoothers are constructed with \texttt{smooth2random()} from \pkg{mgcv} \citep{wood2017a}, similarly to \pkg{gamm4} \citep{wood2020} and \pkg{brms} \citep{brms}.
The statistical model itself is written as a \proglang{C++} template for \TMB.
\TMB\ returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016}.
The negative marginal log likelihood is minimized via the non-linear optimization routine \texttt{stats::nlminb()} \citep{gay1990} in \proglang{R} with additional optimization carried out via a Newton optimization step to find the Hessian using \texttt{stats::optimHess()} \citep{gay1990, r2021}.
After rapid model exploration with maximum likelihood, one can optionally pass an \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017, rstan} via \pkg{tmbstan} \citep{monnahan2018} to samples from the joint posterior distribution for Bayesian inference.

@@

<<dog-binomial-mesh2, results='hide', message=FALSE, warning=FALSE, echo=TRUE, fig.cap="SPDE mesh (lines) combined with the trawl survey observations (points). The locations where lines intersect are referred to as ``vertices''. Finer meshes will be slower to fit but generally increase the accuracy of the SPDE approximation. A greater degree of control over the mesh construction can be achieved by using \\pkg{fmesher} or \\proglang{R}-\\pkg{INLA} directly.", fig.width=5>>=
mesh_pcod2 <- make_mesh(
  pcod, c("X", "Y"),
  fmesher_func = fmesher::fm_mesh_2d_inla,
  cutoff = 8,
  max.edge = c(10, 40),
  offset = c(10, 40)
)
plot(mesh_pcod2)
@

@@

We can generate an area-weighted population index (e.g., a relative or absolute
index of abundance or biomass) that is independent of sampling locations by
predicting from the model on a grid covering the area of interest and summing
the predicted biomass with the \texttt{get\_index()} function (Figure
\ref{fig:dog-index-plot}). We supply the grid cell area (4
km\textsuperscript{2}) to the \texttt{area} argument and specify
\texttt{bias\_correct = TRUE} to enable a bias correction needed due to the
non-linear transformation of the random effects \citep{thorson2016bias}.

% TODO and the Laplace?

@@

\begin{figure}[htb]
\centering
\includegraphics[width=3.5in]{../../figs/timing-spatial-2023-03-16-xkcd.pdf}
\caption[]{Comparison of time to fit an SPDE spatial random field model with
  1000 observations, an intercept and one predictor, Gaussian error, and
  a sequence of SPDE resolutions. Lines represent means and ribbons 95\%
  quantiles across 50 random iterations. Note the log-distributed x axis.
  \pkg{VAST} should be similar to \pkg{sdmTMB} and so is not shown.
  \pkg{inlabru} used the empirical Bayes integration strategy and Gaussian
  approximation with \texttt{bru\_max\_iter = 1}, and the \texttt{like()}
  formulation. \pkg{mgcv} used \texttt{bam()}, \texttt{method = `fREML'}, and
  discretized covariates \citep{miller2019}. Note that \pkg{spaMM} only fits
  spatial, not spatiotemporal, models. All platforms were restricted to one core
and could be faster with parallel computation or optimized algebra
libraries.}\label{fig:timing}
\end{figure}

>>>>

Data are often collected in space or in space repeatedly over time. While such data are a rich source of information across many fields \citep[e.g.,][]{legendre1989a, rossi1992, tilman1997spatial},
% TODO: other non-ecology refs
they are challenging to properly model---data closer in space and time are usually more similar to each other than data farther apart due to measured and unmeasured variables \citep{cressie1993, diggle2007, cressie2011}. 
While measured variables can be accounted for with predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables can cause residual spatial correlation. 
Accounting for this residual correlation is important because doing so allows for valid statistical inference \citep{legendre1989a, dormann2007}, can improve predictions \citep[e.g.,][]{shelton2014}, and can be of interest itself \citep[e.g.,][]{thorson2019d, barnett2021}.

Geostatistical generalized linear mixed effects models (GLMMs) with spatially correlated random effects are a class of models appropriate for these data \citep{rue2005gmrf, diggle2007, cressie2011}. 
Similarly to how random intercepts can account for correlation among groups, spatial or spatiotemporal random effects can account for unmeasured variables causing observations to be correlated in space or space and time. 
A common approach to modelling these spatial effects is with Gaussian random fields (GRFs), where the random effects describing the spatial patterning are assumed to be drawn from a multivariate normal distribution, constrained by some covariance function such as the exponential or Matérn \citep{cressie1993, diggle2007, chiles1999}.
% TODO: decide on using GRF throughout

Such models quickly become computationally limiting due to the need to invert large matrices to keep track of covariation among data. 
Many solutions have been proposed, such as predictive processes \citep{banerjee2008, latimer2009}, the stochastic partial differential equation (SPDE) approximation to GRFs \citep{lindgren2011}, and nearest-neighbour Gaussian processes \citep{datta2016, finley2022}. 
These approaches reduce the scale of the covariance estimation problem while providing a means to evaluate the data likelihood, thereby allowing fitting via Bayesian \citep{gelfand2017} or maximum likelihood methods. 
This can greatly improve computational efficiency \citep[e.g.,][]{heaton2019}. 
The SPDE approach is a solution popularized via the \proglang{R}-\pkg{INLA} package \citep{rue2009, lindgren2011, lindgren2015} and an implementation in \pkg{TMB} \citep[Template Model Builder,][]{kristensen2016} that relies on \proglang{R}-\pkg{INLA} to create input matrices \citep[e.g.,][]{thorson2019, osgood-zimmerman2021}.

% Details are beyond the scope of this paper and are not necessary to use the software discussed here, but the idea is that the solution to a specific SPDE is a GRF with a Matérn covariance function and this `trick' enables one to efficiently fit approximations to GRFs to large spatial datasets \citep{lindgren2011}.

@@

\sdmTMB\ relies on several well-established \R\ packages (Figure~\ref{fig:flowchart}).
\sdmTMB\ uses \pkg{fmesher} \citep{fmesher} and \INLA\ \citep{lindgren2015} to generate the triangulation mesh and input matrices for the SPDE calculations.
The SPDE precision matrix calculations themselves are performed by functions within \pkg{TMB} \citep{kristensen2016} (\texttt{R\_inla} namespace).
Data structures required to implement penalized smoothers are constructed with \texttt{smooth2random()} from \pkg{mgcv} \citep{wood2017a}, similarly to \pkg{gamm4} \citep{wood2020} and \pkg{brms} \citep{brms}.
The statistical model itself is written as a \proglang{C++} template for \TMB.
\TMB\ returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016}.
The negative marginal log likelihood is minimized via the non-linear optimization routine \texttt{stats::nlminb()} \citep{gay1990} in \proglang{R} with additional optimization carried out via a Newton optimization step to find the Hessian using \texttt{stats::optimHess()} \citep{gay1990, r2021}.
After rapid model exploration with maximum likelihood, one can optionally pass an \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017, rstan} via \pkg{tmbstan} \citep{monnahan2018} to samples from the joint posterior distribution for Bayesian inference.
% TODO: visreg + ggeffects effects, etc.?

@@

<<dog-binomial-mesh2, results='hide', message=FALSE, warning=FALSE, echo=TRUE, fig.cap="SPDE mesh (lines) combined with the trawl survey observations (points). The locations where lines intersect are referred to as ``vertices''. Finer meshes will be slower to fit but generally increase the accuracy of the SPDE approximation. A greater degree of control over the mesh construction can be achieved by using \\pkg{fmesher} or \\proglang{R}-\\pkg{INLA} directly.", fig.width=5>>=
# mesh_pcod2 <- make_mesh(
#   pcod, c("X", "Y"),
#   fmesher_func = fmesher::fm_mesh_2d_inla,
#   cutoff = 8,
#   max.edge = c(10, 40),
#   offset = c(10, 40)
# )
# plot(mesh_pcod2)
inla_mesh <- fmesher::fm_mesh_2d_inla(
  loc = cbind(pcod$X, pcod$Y), # coordinates
  max.edge = c(10, 40), # max triangle edge length; inner and outer meshes
  offset = c(10, 40),  # inner and outer border widths
  cutoff = 8 # minimum triangle edge length
)
mesh_pcod2 <- make_mesh(pcod, c("X", "Y"), mesh = inla_mesh)
plot(mesh_pcod2)
@

@@

We can generate an area-weighted population index (e.g., a relative or absolute index of abundance or biomass) that is independent of sampling locations by predicting from the model on a grid covering the area of interest and summing the predicted biomass with the \texttt{get\_index()} function (Figure \ref{fig:dog-index-plot}).
We supply the grid cell area (4 km\textsuperscript{2}) to the \texttt{area} argument and specify \texttt{bias\_correct = TRUE} to enable a bias correction needed due to the non-linear transformation of the random effects \citep{thorson2016bias}.
% TODO and the Laplace?

@@

\begin{figure}[htb]
\centering
\includegraphics[width=3.5in]{../../figs/timing-spatial-2023-03-16-xkcd.pdf}
\caption[]{
Comparison of time to fit an SPDE spatial random field model with 1000 observations, an intercept and one predictor, Gaussian error, and a sequence of SPDE resolutions.
Lines represent means and ribbons 95\% quantiles across 50 random iterations.
Note the log-distributed x axis. 
\pkg{VAST} should be similar to \pkg{sdmTMB} and so is not shown.
\pkg{inlabru} used the empirical Bayes integration strategy and Gaussian approximation with \texttt{bru\_max\_iter = 1}, and the \texttt{like()} formulation.
\pkg{mgcv} used \texttt{bam()}, \texttt{method = `fREML'}, and discretized covariates \citep{miller2019}.
Note that \pkg{spaMM} only fits spatial, not spatiotemporal, models.
All platforms were restricted to one core and could be faster with parallel computation or optimized algebra libraries.}
\label{fig:timing}
\end{figure}
